\chapter{Introduction}

This thesis is about interactive proof development for separation logic.
In the introductory chapter we give an overview of the history of developments of mechanized tools for separation logic, highlight some of the gaps of existing developments and summarize our contributions to the subject.

\section{Separation logic and mechanization}
\label{sec:separation-logic-mechanization}

Separation logic (SL) was introduced by \citet{reynoldsSeparationLogicLogic2002} and
\citet{ohearnLocalReasoningPrograms2001} as a solution to problems with reasoning about mutable data structures.
It proved to be widely successful \cite{ohearnSeparationLogic2019} and served as a basis for multiple other developments of logics used for reasoning about programs, the most influential of them, perhaps, being Concurrent Separation Logic (CSL) by \citet{ohearnResourcesConcurrencyLocal2007}.

In the last decade there were multiple new advanced logics developed that are both higher-order and based on Concurrency Separation Logic, like the one that serves as a foundation for Verified Software Toolchain by \citet{appelProgramLogicsCertified2014} or Iris \cite{jungIrisGroundModular2018}, which was used to establish formal foundations of the Rust programming language \cite{jungRustBeltSecuringFoundations2018}.

Separation logic also gave rise to multiple mechanized tools.
Initial applications of separation logic to mechanized developments were concentrating on mostly automatic proofs, like Smallfoot~\cite{berdineSmallfootModularAutomatic2006} or Verifast~\cite{jacobsVeriFastProgramVerifier2008} or Bedrock~\cite{chlipalaMostlyautomatedVerificationLowlevel2011, chlipalaBedrockStructuredProgramming2013}.
With this approach logics have to be restricted in order for automation to be feasible, as on its own even separation logic with a magic wand \((\wand)\) is undecidable \cite{brotherstonUndecidabilityPropositionalSeparation2014}.

However, automatic proofs are not the only possibility.
They are also not suitable for more expressive variants of SL, which can include advanced connectives, such as modalities and views.
The alternative to automatic verification is tactic-based interactive verification.
There are multiple developments that pursue this idea, usually utilizing existing interactive proof assistants: ``Charge!''~\cite{bengtsonCharge2012}, VST-Floyd~\cite{caoVSTFloydSeparationLogic2018} for Verified Software Toolchain, and MoSeL for Iris~\cite{krebbersInteractiveProofsHigherorder2017, krebbersMoSeLGeneralExtensible2018}.
{
\interfootnotelinepenalty=10000
In fact, MoSeL (and its predecessor, IPM -- Iris Proof Mode \footnote{We are going to focus on MoSeL for the rest of the thesis. The title of the thesis stays true to the nature of the work, since the biggest application of MoSeL is still proof mode for Iris.}) turned out to be essential for developments based on Iris \cite{krebbersMoSeLGeneralExtensible2018, jungUnderstandingEvolvingRust2020}: verification of Rust's safety claims \cite{jungRustBeltSecuringFoundations2018, dangRustBeltMeetsRelaxed2019}, formalization of Scala's core type system \cite{giarrussoScalaStepbystepSoundness2020}, a framework for concurrent reasoning based on message-passing \cite{hinrichsenActrisSessiontypeBased2019} and many others.}

VST-Floyd~\cite{caoVSTFloydSeparationLogic2018} and MoSeL \cite{krebbersMoSeLGeneralExtensible2018} are two most recent major contributions in the direction of tactic-based interactive verification.
Both of them provide support for automation, though they differ significantly in approaches.
VST-Floyd follows the idea of normal form for goals, which was also used in \cite{bengtsonCharge2012} and is focused on tactics for one specific logic and forward-style reasoning approach.
MoSeL, on the other hand, doesn't try to normalize goals and places emphasis on genericity of the logic: it builds on the interface of Modal Bunched Implications (MoBI) logic, which allows the user to apply it in any logic that instantiates the interface.

\section{Problems and approaches}
\label{sec:problems-approaches-intro}

While MoSeL achieved a lot, there still are some issues with the current state of the tool.
We now describe the problems and our approach to solving them in this thesis.

\subsection{Ltac1 issues}
\label{sec:ltac1-issues}
Ltac was the first tactic language for Coq that allowed extensive automation and it served Coq well, for which it was hailed as a ``major ingredient of the success'' of Coq \cite{pedrotLtac2TacticalWarfare2019}.
Unfortunately, it has outgrown its original purpose and is now known to be problematic for developments of non-trivial complexity.
It has badly defined semantics, such that even maintainers don't understand the order of evaluation \cite{pedrotLtacInternals2016}, lacks clear distinction between Coq variables and Ltac variables \cite{pedrotLtacInternals2016}, proper exceptions, the ability to deal with existential variables.
Even the documentation says that some tactics should be used ``at your own risk'' \cite[Section~3.3.1]{thecoqdevelopmentteamCoqProofAssistant2020}.
MoSeL also relies on Ltac to a great extent and this is a handicap for its development and maintenance, because of the fragility of larger programs written in Ltac.

Fortunately, since the problems with Ltac are well-known, many other languages appeared to substitute it \cite{malechaExtensibleEfficientAutomation2016, zilianiMtacMonadTyped2013, kaiserMtac2TypedTactics2018a, tassiElpiExtensionLanguage2018}.
There is also a new experimental tactic language that aims to be a direct successor of Ltac, tries to stay as close as possible to Ltac, while maintaining reasonable semantics: Ltac2 \cite{pedrotLtac2TacticalWarfare2019}.

In an attempt to evaluate Ltac2 as a language for bigger developments, we rewrite Ltac parts of MoSeL to Ltac2.
We tried to improve the implementation where possible and use this new version of MoSeL to develop new features.

\subsection{The \texttt{iSplit} tactic}
\label{sec:isplit-tactic}

One of the main problems with incorporating backwards-style proofs into separation logic is the introduction rule for a new connective: ``separating conjunction''.
\[\infer{\Gamma_1, \Gamma_2 \vdash \phi_1 \ast \phi_2}
        {\Gamma_1 \vdash \phi_1 &
         \Gamma_2 \vdash \phi_2} \]
The logic is substructural since it lacks the contraction rule, so introduction of separating conjunction forces us to distribute hypotheses between two branches of the proof.
However, this creates a problem with bottom-up proof construction, since we frequently don't know which propositions will be needed in which branch until it's finished.

In MoSeL, introduction of separating conjunction corresponds to the \coqe{iSplit} tactic (separation logic variant of \coqe{split}).
It requires the user to provide a list of hypotheses that will be attributed to either the left (\coqe{iSplitL}) or the right (\coqe{iSplitR}) branch, leaving it for the user to essentially guess beforehand which hypotheses are going to be used where.

The issue with the rule was first described in a paper by \citet{harlandResourceDistributionBooleanConstraints2003}.
They proposed an amended entailment predicate to solve the problem.
The idea is to postpone the distribution by attaching Boolean flags (\(e_1, e_2\)) that indicate presence of hypotheses and subsequently imposing the side-condition that the flags are disjoint.
\[\infer[e_i \oplus \hat{e}_i = \true, i \in \{1,2\}]
        {P, Q \vdash \phi_1 \ast \phi_2}
        {P[e_1], Q[e_2] \vdash \phi_1 &
         P[\hat{e}_1], Q[\hat{e}_2] \vdash \phi_2} \]
We adapt their solution to MoSeL and create a new \coqe{iSplit} tactic which doesn't require the user to provide such a list.

\subsection{Goal matching}
\label{sec:goal-matching}

A very useful tactical of Ltac is \coqe{match goal}.
It allows the user to match the hypothesis and the goal in the current proof state and has proven to be essential for automation tactics.
E.g.\ it is used in \coqe{naive_solver} from stdpp~\cite{std++developersandcontributorsStdpp}, one of the first examples in CPDT~\cite{chlipalaCertifiedProgrammingDependent2013} and in the \coqe{break_match} tactic from StuctTact library~\cite{uwplse-structtactdevelopmentteamStructTact2020}.

\begin{figure}[H]
\begin{coq}
match goal with
| [H : False |- _] => exfalso; exact H
| [|- True] => exact I
end.
\end{coq}
  \caption{Small example of automation tactic which uses \coqe{match goal}}
  \label{fig:match-goal-example-intro}
\end{figure}

MoSeL doesn't use Coq contexts and instead stores hypotheses in the goal, which means that the original tactical is of limited usage, and Ltac doesn't provide an obvious opportunity to implement goal-matching for MoSeL contexts, i.e.\ there is no way to reuse the syntax of goal matching for user-defined tactics.
We develop a new tactic \coqe{iMatch goal} in Ltac2 for the reimplemented MoSeL, which aims to be a separation-logic alternative to \coqe{match goal}.

\section{Contributions}
\label{sec:contributions}

\begin{itemize}
\item In chapter~\ref{chap:reimplementation_ipm}, we describe our reimplementaion of MoSeL in Ltac2, which is, to our knowledge, contributing one of the first significant applications of the language.
  In the process, we evaluate the parts of Ltac2 used and suggest possible directions for future development.
\item In chapter~\ref{chap:postponed_splitting}, we develop an adaptation of results by \citet{harlandResourceDistributionBooleanConstraints2003} to MoSeL, as reimplemented in the previous chapter, to introduce a new splitting tactic \coqe{iSplit}.
\item In chapter~\ref{chap:imatch}, we describe a new primitive \coqe{iMatch goal} for the tactic language of separation logic which allows us to match entailments in a fashion similar to Coq's native \coqe{match goal} construct.
\item Finally, in chapter~\ref{chap:solver} we utilize the previously implemented features in order to implement a simple solver for separation logic.
\item In chapter~\ref{cha:related-work} we give an overview of related work and list some possible directions for future work.
\end{itemize}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "thesis"
%%% TeX-parse-self: t
%%% TeX-auto-save: t
%%% reftex-cite-format: natbib
%%% reftex-default-bibliography: ("/home/buzzer/my-dir/ed/uni/saar/prjcts/iris/npm/tex/TacticsProofs.bib")
%%% End: